Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-01-06 paper/tidy-missing-data-paper.Rmd
reference example data in text
add inline code into tables

```{r gg-box-na, fig.height = 6, fig.width = 6, out.width = "70%", fig.align = "center", fig.cap = "ggplot2 provides different visualizations depending on what type of data has missing values. (A) Data is complete and graphic is presented; (B) Missings are only in category variable year - an additional 'NA' boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and school year, additional missing information is shown. The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)). In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.", fig.show = "asis"}

@@

```{r vis-miss, fig.height = 4, fig.width = 6, out.width = "95%", fig.show = 'hold', echo = FALSE, dev = "png", dpi = 300, fig.cap = "Heatmap visualizations of missing data for the airquality dataset. (A) The default output and (B) ordered by clustering on rows and columns. There are only missings in ozone and solar radiation, and there appears to be some structure to their missingness."}

@@

miss_var_summary(airquality) %>% 
  knitr::kable(
    caption = "Summary of the number and percent of missings in each variable. Only ozone and solar radiation have missing values.",
    digits = 1)
    # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))

@@

miss_var_table(airquality) %>% 
  knitr::kable(
    caption = "Tabulation of the amount of missing data in each variable. This shows the number of variables with no missings, 7 missings, and 37 missings, and the percentage of variables with those amounts of missingness. There are not many patterns of missingness.",
    digits = 1)
    # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))
```

@@

airquality %>%
  group_by(Month) %>%
  miss_var_summary() %>%
  ungroup() %>%
  slice(1:10) %>%
  knitr::kable(caption = "A summary of the missingness in each variable, for each Month of the airquality dataset, with only the first 10 rows of output shown. There are more ozone missings in June than May.",
               digits = 1)
               # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))

@@

There are many common values that mean "missing". For example, "N/A" "Not Available", "MISSING", or white space " ". Before replacing these values with missing values (`NA`), it is useful to first _scan_ the data to get a sense of the magnitude of the problem. The  `miss_scan_count()` function finds where specified values occur in data. For example, for the example data, `dat_ms`:

@@

All occurrences of -99 can be searched for using the code:

@@

@@

`naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation, such as `mice`, `mi`, `VIM`, and `simputation`. However, `naniar` does provide a few imputation methods to facilitate exploration and visualizations, which were not otherwise available: `impute_below` and `impute_mean` and `impute_median`.  These imputation methods are useful to explore structure in missingness, but are not recommended for use in analysis.

The `impute_below` function imputes values below the minimum of the data, with some jitter to reduce overplotting, but has options to change the default amount it is shifted below, and the amount of jitter.

```{r demonstrate-impute-below}
airquality %>%
  impute_below_at(vars(Ozone)) %>%
  select(Ozone, Solar.R) %>%
  head()
```

@@

One drawback to imputation functions in this form is that the location of imputed values is not tracked. This is covered in Section \@ref(verbs-track).

@@

aq_imputed <- airquality %>%
  bind_shadow() %>%
  as.data.frame() %>%
  simputation::impute_knn(Ozone ~ .) %>%
  simputation::impute_knn(Solar.R ~ .) %>%
  add_label_shadow() %>%
  as_tibble()

Missing values can then be shown in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot looking at one variable, using the `fill = any_missing`, (Figures \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C).   
```r
ggplot(aq_imputed,
       aes(x = Ozone,
           y = Solar.R,
           color = any_missing)) + 
  geom_point() +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "bottom")

ggplot(aq_imputed,
       aes(x = Ozone,
           fill = any_missing)) + 
  geom_density(alpha = 0.3) + 
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "none")

ggplot(aq_imputed,
       aes(x = Solar.R,
           fill = any_missing)) + 
  geom_density(alpha = 0.3) + 
  scale_fill_brewer(palette = "Dark2")
```

```{r track-impute-example, fig.show = "hold", fig.cap = "Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values imputed using simputations `impute\\_knn` function, with imputed values colored green and data values orange. Imputed values are similar, but slightly different to existing data.", fig.height = 2, fig.width = 5, out.width = "100%", echo = FALSE}

ggplot(aq_imputed,
       aes(x = Ozone,
           y = Solar.R,
           color = any_missing)) + 
  geom_point() +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none") + 
  labs(tag = "A")

ggplot(aq_imputed,
       aes(x = Ozone,
           fill = any_missing)) + 
  geom_density(alpha = 0.3) + 
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "none") + 
  labs(tag = "B")

ggplot(aq_imputed,
       aes(x = Solar.R,
           fill = any_missing)) + 
  geom_density(alpha = 0.3) + 
  scale_fill_brewer(palette = "Dark2") + 
  theme(legend.position = "none") + 
  labs(tag = "C")

@@

```r
aq_imputed %>%
  group_by(any_missing) %>%
  summarise_at(.vars = vars(Ozone),
               .funs = list(min, mean, median, max))
```

@@

```{r applic-vis-miss, fig.height = 5, fig.width = 6.25, out.width = "85%", fig.show='hold', fig.cap = "Heatmap of clustered missingness for the housing data. Three groups of missingness are apparent, at the top for building area to longitude, the middle for building area and year built, at the end for building area, year built, and landsize. There is some structure in the missings.", dev = "png", dpi = 300}

@@

The new data structures discussed in the paper could be used to create different visualizations than were shown in the paper. The analyst can use the data structures to decide on appropriate visualization for their problem. The data structures could also be used to support interactive graphics, in the manner of MANET and ggobi. Linking the plots (linked brushing) to explore missingness or animating between different sets of imputed values. New packages like `plotly` [@plotly] facilitate interactive graphics, and packages like `gganimate` facilitate animations [@gganimate].

>>>>

```{r gg-box-na, fig.height = 3, fig.width = 9, out.width = "100%", fig.align = "center", fig.cap = "ggplot2 provides different visualizations depending on what type of data has missing values. (A) Data is complete and graphic is presented; (B) Missings are only in category variable year - an additional 'NA' boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and school year, additional missing information is shown. The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)). In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.", fig.show = "asis"}

@@

```{r vis-miss, fig.height = 3, fig.width = 4.5, out.width = "80%", fig.show = 'hold', echo = FALSE, dev = "png", dpi = 300, fig.cap = "Heatmap visualizations of missing data for the airquality dataset. (A) The default output and (B) ordered by clustering on rows and columns. There are only missings in ozone and solar radiation, and there appears to be some structure to their missingness."}

@@

miss_var_summary(airquality) %>% 
  knitr::kable(
    caption = "The \\texttt{miss\\char`_var\\char`_summary} function provides a summary of the number and percent of missings in each variable. Only ozone and solar radiation have missing values.",
    digits = 1)
    # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))

@@

miss_var_table(airquality) %>% 
  knitr::kable(
    caption = "The \\texttt{miss\\char`_var\\char`_table} function provides a tabulation of the amount of missing data in each variable. This shows the number of variables with no missings, 7 missings, and 37 missings, and the percentage of variables with those amounts of missingness. There are not many patterns of missingness.",
    digits = 1)
    # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))
```

@@

airquality %>%
  group_by(Month) %>%
  miss_var_summary() %>%
  ungroup() %>%
  slice(1:10) %>%
  knitr::kable(caption = "The \\texttt{miss\\char`_var\\char`_summary} provides a grouped summary of the missingness in each variable, for each Month of the airquality dataset, with only the first 10 rows of output shown. There are more ozone missings in June than May.",
               digits = 1)
               # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))

@@

There are many common values that mean "missing". For example, "N/A" "Not Available", "MISSING", or white space " ". Before replacing these values with missing values (`NA`), it is useful to first _scan_ the data to get a sense of the magnitude of the problem. The  `miss_scan_count()` function finds where specified values occur in data. For example, for the example data, `dat_ms`  Table \@ref(tab:setup-miss-scan-count). All occurrences of -99 can be searched for using the code:

@@

@@

(**considering removing these examples**)

@@

`naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation, such as `mice`, `mi`, `VIM`, and `simputation`. However, `naniar` does provide a few imputation methods to facilitate exploration and visualizations, which were not otherwise easily available: `impute_below` and `impute_mean` and `impute_median`.  These imputation methods are useful to explore structure in missingness, but are not recommended for use in analysis.

The `impute_below` function imputes values below the minimum of the data, with some jitter (small values of random noise) to reduce overplotting, but has options to change the default amount it is shifted below, and the amount of jitter.

```{r demonstrate-impute-below, eval = FALSE}
airquality %>%
  impute_below_at(vars(Ozone)) %>%
  select(Ozone, Solar.R) %>%
  head()
```

@@

One drawback to imputation functions in this form is that the location of imputed values is not tracked. This issue is resolved with \code{nabular} data covered in Section \@ref(verbs-track).

@@

aq_imputed <- airquality %>%
  bind_shadow() %>%
  simputation::impute_lm(Ozone ~ Temp + Wind) %>%
  simputation::impute_lm(Solar.R ~ Temp + Wind) %>%
  add_label_shadow()

aq_imputed
```

Missing values can then be shown in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot looking at one variable, using the `fill = any_missing`, (Figures \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C).

```{r track-impute-example, fig.show = "hold", fig.cap = "Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values imputed using simputations `impute\\_knn` function, with imputed values colored green and data values orange. Imputed values are similar, but slightly different to existing data.", fig.height = 4, fig.width = 12, out.width = "100%", echo = FALSE}

p1 <- 
ggplot(aq_imputed,
       aes(x = Ozone,
           y = Solar.R,
           color = any_missing)) + 
  geom_point() +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none") + 
  labs(tag = "A")

p2 <- 
ggplot(aq_imputed,
       aes(x = Ozone,
           fill = any_missing)) + 
  geom_density(alpha = 0.3) + 
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "none") + 
  labs(tag = "B")

p3 <- 
ggplot(aq_imputed,
       aes(x = Solar.R,
           fill = any_missing)) + 
  geom_density(alpha = 0.3) + 
  scale_fill_brewer(palette = "Dark2") + 
  theme(legend.position = "none") + 
  labs(tag = "C")

gridExtra::grid.arrange(p1, p2, p3, nrow = 1)
```

@@

```r
aq_imputed %>%
  group_by(any_missing) %>%
  summarise_at(.vars = vars(Ozone),
               .funs = lst(min, mean, median, max))
```

@@

```{r applic-vis-miss, fig.height = 4, fig.width = 6, out.width = "85%", fig.show='hold', fig.cap = "Heatmap of clustered missingness for the housing data. Three groups of missingness are apparent, at the top for building area to longitude, the middle for building area and year built, at the end for building area, year built, and landsize. There is some structure in the missings.", dev = "png", dpi = 300}

@@

The new data structures discussed in the paper promote flexibility, and could be used to create different visualizations than were shown in the paper. The analyst can use the data structures to decide on appropriate visualization for their problem. The data structures could also be used to support interactive graphics, in the manner of MANET and ggobi. Linking the plots (linked brushing) to explore missingness or animating between different sets of imputed values. New packages like `plotly` [@plotly] facilitate interactive graphics, and packages like `gganimate` facilitate animations [@gganimate].
